Peaks Over Threshold for Bursty Time Series

### 2018-04-20 article_springer/article_springer.Rmd

As will be shown below, in the heavy-tailed waiting time scenario threshold 
crossing times form a _fractional Poisson process_ 
[@Laskin2003; @Meerschaert2010b], which is a 
renewal process with Mittag-Leffler distributed waiting times. 
The family of Mittag-Leffler distributions nests the exponential 
distribution [@Haubold11], and hence the fractional Poisson process
generalizes the standard Poisson process. 
Again as the threshold size increases and the threshold crossing
probability $p$ decreases, the fractional Poisson process is rarefied: 
The scale parameter of the Mittag-Leffler inter-arrival times of 
threshold crossing times increases, increases, but _superlinearly_; 
see the Theorem below.

@@

Without loss of generality, we choose $\ell^*$ such that $F(\ell^*) = 1/e$, 
and let $\ell = a(\tilde b(c)) \ell^* + d(\tilde b(c))$. 
We may then calculate 
$$
\mathbf P\left[ \frac{T(\ell)}{b(1/p)} > t \right]
= \mathbf P[T(\ell) > b(1/p) t]
= \mathbf P[V(ct) \le \ell]
= \ldots
$$
where we have substituted $c = b(1/p)$, 
$$
\ldots
= \mathbf P\left[ \frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\le \frac{\ell - d(\tilde b(c))}{a(\tilde b(c))} \right]
\longrightarrow \mathbf P\left[ Y(t) \le \ell^* \right]
= \ldots
$$
Defining the hitting time of level $\ell^*$ by $Y(t)$ as 
$\xi_{\ell^*} := \inf\{t: Y(t) > \ell^*\}$,
we then have 
$$
\ldots = \mathbf P[\xi_{\ell^*} > t] 
= \mathbf P[(-\log F(\ell^*))^{-1/\beta} X^{1/\beta} D > t]
$$
by [Proposition 4.2, @MeerschaertStoev08], where $X$ is an exponential 
random variable with mean $1$. 
Using [Theorem 19.1, @Haubold11], we see that 
$X^{1/\beta} D \sim {\rm ML}(\beta, 1)$, concluding the proof. 
\qed

@@

I.i.d.:
: After removing the "noise observations" below the smallest threshold 
  $\ell_0$, the pair sequence $(T(\ell_0, i), X(\ell_0,i))$ is i.i.d. 
  An indication if this is true is given 
  by an auto-correlation plot for the logarithms (to ensure finite moments) 
  of the two time series.

@@

The "anomalous" scaling behaviour in the bursty setting entails two 
phenomena: 
i) a heavy tail of the interarrival time distribution of 
threshold crossings (long rests), and 
ii) a high propensity for more threshold crossing events immediately after 
each threshold crossing event (bursts). 
The Mittag-Leffler distribution captures both phenomena, due to its heavy
tail as well as its streteched exponential (peaked) asymptotics for small 
times. 
It generalizes the exponential distribution, and 
in the solar flare data example, this generalization 
is warranted, because the likelihood-ratio test is strongly significant.

>>>>

As will be shown below, in the heavy-tailed waiting time scenario threshold 
crossing times form a _fractional Poisson process_ 
[@Laskin2003; @Meerschaert2010b], which is a 
renewal process with Mittag-Leffler distributed waiting times. 
The family of Mittag-Leffler distributions nests the exponential 
distribution [@Haubold11], and hence the fractional Poisson process
generalizes the standard Poisson process. 
Again as the threshold size increases and the threshold crossing
probability $p$ decreases, the fractional Poisson process is rarefied: 
The scale parameter of the Mittag-Leffler inter-arrival times of 
threshold crossing times increases, but _superlinearly_; 
see the Theorem below.

@@

Without loss of generality, we choose $\ell^*$ such that $F(\ell^*) = 1/e$, 
and let $\ell = a(\tilde b(c)) \ell^* + d(\tilde b(c))$. 
We may then calculate 
$$
\mathbf P\left[ \frac{T(\ell)}{b(1/p)} > t \right]
= \mathbf P[T(\ell) > b(1/p) t]
= \mathbf P[V(ct) \le \ell]
$$
where we have substituted $c = b(1/p)$. Moreover 
$$
\mathbf P[V(ct) \le \ell]
= \mathbf P\left[ \frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\le \frac{\ell - d(\tilde b(c))}{a(\tilde b(c))} \right]
\longrightarrow \mathbf P\left[ Y(t) \le \ell^* \right]
$$
Defining the hitting time of level $\ell^*$ by $Y(t)$ as 
$\xi_{\ell^*} := \inf\{t: Y(t) > \ell^*\}$,
we then have 
$$
P\left[ Y(t) \le \ell^* \right] = \mathbf P[\xi_{\ell^*} > t] 
= \mathbf P[(-\log F(\ell^*))^{-1/\beta} X^{1/\beta} D > t]
$$
by [Proposition 4.2, @MeerschaertStoev08], where $X$ is an exponential 
random variable with mean $1$. 
Using [Theorem 19.1, @Haubold11], we see that 
$X^{1/\beta} D \sim {\rm ML}(\beta, 1)$, concluding the proof. 
\qed

@@

i.i.d.:
: After removing the "noise observations" below the smallest threshold 
  $\ell_0$, the pair sequence $(T(\ell_0, i), X(\ell_0,i))$ is i.i.d. 
  An indication if this is true is given 
  by an auto-correlation plot for the logarithms (to ensure finite moments) 
  of the two time series.

@@

The "anomalous" scaling behaviour in the bursty setting entails two 
phenomena: 
i) a heavy tail of the interarrival time distribution of 
threshold crossings (long rests), and 
ii) a high propensity for more threshold crossing events immediately after 
each threshold crossing event (bursts). 
The Mittag-Leffler distribution captures both phenomena, due to its heavy
tail as well as its stretched exponential (peaked) asymptotics for small 
times. 
It generalizes the exponential distribution, and 
in the solar flare data example, this generalization 
is warranted, because the likelihood-ratio test is strongly significant.
